---
title: "DATA 202 - Week 1"
subtitle: "Introduction to the Course"
author: "Nathan Alexander, PhD"
institute: "Center for Applied Data Science and Analytics"
bibliography: references.bib # include references file
csl: apa.csl
format: 
  html: default
  pptx:
    toc: false
    reference_docx: "word-styles-reference.docx"
    fig_caption: true
  revealjs:
    output-file: week01-slides.html
    height: 900
    width: 1600
    smaller: false
    slide-number: c/t #< collapsted/total
    logo: "img/howard-logo.jpg"
    footer: "[Course Data GitHub](https://github.com/data-202)"
    toc: false
    echo: true
    incremental: false
geometry: margin=1.0in
editor_options:
  markdown:
    wrap: sentence
---

```{r, echo=F, eval=F, warning=FALSE}
install.packages("devtools")
library(devtools)
```

```{r, include=F, warning=F}
library(tidyverse) #collection of R packages designed for data science
library(here) #helps with filepaths
library(gt) # creates tables
library(DiagrammeR) # create path diagrams
library(data.table) # use for `melt()` function
here::i_am("week01.qmd")
```

## Course Description

In this course, students will develop an understanding of statistics as a research tool.
Students are expected to have some basic knowledge of statistics from a prior course.
Emphasis will be placed on understanding statistical concepts and applying and interpreting tests of statistical inference for real-life applications.
The content will include, but not be limited to, visual representations of data, descriptive statistics, correlation and simple regression, sampling distributions, and the assumptions associated with and the application of selected inferential statistical procedures.
Throughout the course, there will be a strong emphasis on how statistical modeling can be a driving force for social justice.

## Course Learning Objectives

-   Appreciate and understand the role of statistics in your field

-   Evaluate, comprehend, and explain the statistical findings in a data set

-   Explore data in R

-   Apply appropriate application and interpretation of various inferential statistical procedures

-   Write a simple description of methodology and results from analysis

-   Develop an ability to apply appropriate statistical methods to summarize and analyze data

-   Make sense of data and be able to report the results in appropriate tables or statistical terms

## Course companion site

The landing page for our companion site can be found [here](https://data-202.github.io/sp25).

## Initial assignments

### Annotated Bibliography

Your annotated bibliography is due Sun Feb 2 using OpenAlex or the Web of Science (WoS).

### Lab #1

Lab 1 is due Sun Feb 16.

### Paper #1

You can read more about paper #1 [here](/papers/paper1.qmd).

See the instructions for paper assignments [here](/papers/papers-instructions.qmd).

Also, learn more about paper assignments [here](/papers/papers-more-info.qmd).

## Case Study 1: Tuskegee Experiement of Untreated Syphillis

To help you prepare for our forthcoming discussions and readings, you should explore information about our first case study.
One place to start is here: "The Tuskegee Experiment: Crash Course Black American History #29" in the video below:

<center>

<iframe width="560" height="315" src="https://www.youtube.com/embed/56gqCXlUCoE" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen>

</iframe>

</center>
