---
title: "Lab 1"
subtitle: "DATA 202 - Alexander"
format: 
  html:
    self-contained: true
    self-contained-math: true
editor: visual
---

For Lab 1, you should submit your solutions to Canvas as a *RMarkdown* syntax (i.e., you can render an html file or push your code to GitHub, or share the .Rmd file). Please see the *Preparing Lab Reports* section at the bottom of this assignment.

The `Lab 1` assignment will reintroduce you to data in the `critstats` package and add a second data set known as `gss_cat`. However, for the lab 1 and paper 2 assignment, you may use your own data set and we can work together to find you a data set that is ready to be analyzed. For new users, I would recommend against data that needs to be cleaned. However, by visiting the case study 2 files, you can learn more about cleaning your data.

That said, during this lab, you will also return to some of the functions we have used before to clean variables and manipulate data frames. Your analyses in this lab should focus on a bivariate analysis between two categorical variables, two numeric variables, or one numeric variable and one categorical variable.

------------------------------------------------------------------------

## Learning Activities

By the end of this lab you will be able to:

1.  Locate data sets in the `critstats` package
2.  Develop an original research question
3.  Clean and manipulate data for analysis
4.  Examine the relationship between two variables

You should submit your final output on Canvas.

------------------------------------------------------------------------

# Part 0: Pre-lab tasks

-   Create a new RMarkdown

-   Check your working directory

-   Write a preamble

-   Install and/or update packages and load libraries

------------------------------------------------------------------------

## Task 0.1: Create a new RMarkdown

In your R session, navigate to: File \> New File \> R Markdown. Create a new markdown file using an appropriate title.

## Task 0.2: Check your working directory

Check your working directory by typing `getwd()`.

If you are not in the desired directory, you can change your directory using the associated path. This path should be the same as the project folder that you plan to work out of for the next several weeks.

```{r}
#| echo: true
#| output: false
#| warning: false
# insert your desired path in the parenthesis and remove the #
# setwd("/your/working/directory/goes/here") 
```

You can add a new sub-folder manually or under the **Files** tab in the RStudio IDE.

## Task 0.3: Write a preamble

Name: <include your full name> Assignment: Lab 2 Date: <here you may want to add a date> Purpose: <insert the goals or purpose of the RScript>

## Task 0.4: Packages and libraries

```{r}
#| echo: true
#| output: false
#| warning: false
# install the tidyverse package
install.packages("tidyverse", repos = "http://cran.us.r-project.org")

# load the libraries needed for today's analyses
library(tidyverse)
library(dplyr)
library(critstats)

## update packages if needed; remove # to run code
# update.packages("package-name")
```

------------------------------------------------------------------------

# Part 1:

In this part of the lab, you should complete the following tasks:

-   Locate a data set in the `critstats` package or your own data set.

```{r}
library(critstats)
data(package = 'critstats')
```

-   Recall that you can also use the `data(package="critstats")` command, with a double asterisk, to see a list of data sets in the \`critstats' package

-   Develop a set of research questions about two variables in your selected data set.

------------------------------------------------------------------------

# Part 2:

In this part of the lab, you should complete the following tasks:

-   Clean and manipulate your data for analysis

-   Examine the relationship between two variables

------------------------------------------------------------------------

# Part 1 Reports

## Report 1.1

What data set have you decided to use?

## Report 1.2

Which two variables from your data set will be analyzed?

## Report 1.3

What is your research question?

## Report 1.4

What is your data analysis plan? Please be descriptive.

## Report 1.5

What are some potential limitations for your analysis?

------------------------------------------------------------------------

# Part 2 Reports

## Report 1.6

Does your data contain missing values? If so, how have you dealt with these values?

## Report 1.7

Please include all code used to clean and manipulate the variables.

## Report 1.8

What relationship, if any, exists between the two variables?

## Report 1.9

How do these findings relate to your research question and theory?

## Report 1.10

What limitations exist as a result of the data analysis?

# Experiencing issues?

If you experience issues executing your code, it is likely that you have made an error in the previous steps.

Remember that R is case sensitive in all instances, and space sensitive in some instances.

Please be sure to go back and carefully check your code.
