---
title: "Untitled"
author: "Nathan Alexander"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=F, message=F}
# load libraries for this session
library(tidyverse)
library(dplyr)
library(remotes)
```

Load the `critstats` package.

```{r, warning=F, message=F}
# use the remote install function to call in your data
# remotes::install_github("professornaite/critstats", force=TRUE)

# load the critstats library
library(critstats)
```

Load the `true_size` data frame in the `critstats` package.

```{r, warning=F, message=F}
critstats::true_size
```

Label my `true_size` data frame as `df1`.

```{r, warning=F, message=F}
df1 <- critstats::true_size
df1
```

Inspect the data.

```{r, warning=F, message=F}
str(df1) # This will show us the structure of data frame 1 (df1)

dim(df1) # This will show me the dimensions of the data frame

glimpse(df1) # This will give you an output that is similar to str(df1) but it doesn't have as much info
```
View the top part and bottom part of our data frame.

```{r, warning=F, message=F}
head(df1)
tail(df1)

head(df1, n = 10)
tail(df1, n = 10)
```

View a summary of the data frame.

```{r, warning=F, message=F}
summary(df1)
```

We'll look at some of the dplyr variables.

```{r, warning=F, message=F}

select(df1, Country, percent.africa)
select(df1, -area.sq.km)
```

We use `filter` to select row, or deselect rows in our data frame.

```{r}
filter(df1, percent.africa > 30)
filter(df1, percent.africa < 1)
filter(df1, percent.africa < 11 & percent.africa > 0)
```

```{r}
df1
filter(df1, Country == "China")
```

Add and remove columns from a data set.

```{r}
mutate(df1, est.square.miles = df1$area.sq.mi * 1000000)

# let's use a pipe %>% to run the same operation as above
# the pipe operator literally means "and then"
df1 %>%
  mutate(est.square.miles = df1$area.sq.mi * 1000000)
```

Add two new variables and remove the old variables.

```{r}

df1 %>% 
  mutate(est.square.miles = df1$area.sq.mi * 1000000) %>% 
  mutate(est.square.km = df1$area.sq.km * 1000000) %>% 
  select(-area.sq.mi, -area.sq.km) # remove the columns that we do not need
```


```{r}
df1 %>% 
  mutate(est.square.miles = df1$area.sq.mi * 1000000) %>% 
  mutate(est.square.km = df1$area.sq.km * 1000000) %>% 
  select(-area.sq.mi, -area.sq.km) %>% # remove the columns that we do not need
  rename(country = Country) %>% # making the 'C' in the old Country column lowercase
  relocate(country, percent.africa, est.square.km)
```

Overwrite the data frame with all of the changes that I have made.

```{r}
true_size_modified <- df1 %>% 
  mutate(est.square.miles = df1$area.sq.mi * 1000000) %>% 
  mutate(est.square.km = df1$area.sq.km * 1000000) %>% 
  select(-area.sq.mi, -area.sq.km) %>% # remove the columns that we do not need
  rename(country = Country) %>% # making the 'C' in the old Country column lowercase
  relocate(country, percent.africa, est.square.km)

true_size_modified
```








